Manuscript Figures

### 2021-03-25 src/manuscript_figures.Rmd
replace AMI computation, reduce parallel threads to 10

---
title: "Manuscript Figures"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
# todo: pass this as argument
settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/results/hallmark_genes/settings.yaml')
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
library(ggridges)
library(ggcorrplot)
library(ComplexHeatmap)
library(pheatmap)
theme_set(ggpubr::theme_pubclean())
geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
FigureFolder <- '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas/test'

@@

TOOLS <- c('pca_factors', 'maui_factors', 'mofa_factors', 'mcia_factors')

@@

# do unsupervised clustering and compute AMI 
experiment <- 'pancancer.cnv_gex_meth_mut'
df <- do.call(rbind, pbapply::pblapply(TOOLS, function(x) {
  message(x)
  M <- LFs[[x]][[experiment]]
  if(x == 'maui_factors') {
    #M <- remove_redundant_variables(M, perc = 99)
    M <- scale(M)
  }
  # find cancer types per sample
  ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
  data.frame('tool' = x, 
             'AMI' = compute_AMI(M, k = length(unique(ct)), target_labels = ct))
}))
df$tool <- toupper(gsub("_factors", "", df$tool))
p5 <- ggplot(df, aes(x = tool, y = AMI)) +
  geom_bar(aes(fill = tool), stat = 'identity') + 
  geom_text(aes(label = round(AMI, 2))) + 
  theme(legend.title = element_blank(), axis.title.x = element_blank())

@@

multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
cluster.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  stats <- do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, 
                                            intersect(names(LFs[[tool]]), experiments), 
                    function(experiment) {
    message(experiment)
    source(settings$utility_script)
    library(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      #M <- remove_redundant_variables(M, perc = 99) 
      M <- scale(M)
    }

@@

# survival prediction accuracy benchmarking
# find survival-predictive LFs by univariate cox models and use them in RSF
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings', 'surv', 'clin_factors'))
surv.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
  message("survival benchmark for ",tool)
  do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source(settings$utility_script)
    require(data.table)
    require(survminer)
    M <- LFs[[tool]][[experiment]]
    clin <- process_clinical_covariates(rownames(M))
    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
    # compute pfi stats
    # fit a first model to filter variables 
    fit <- fit_RSF(surv, M, ntree = 500, 'PFI.time', 'PFI')
    # using a range of top variables, compute c index
    dt1 <- do.call(rbind, lapply(2:10, function(i) {
      vars <- names(head(sort(fit$importance, decreasing = T), i))
      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'PFI', 'PFI.time'), 
                                M[,vars], ntree = 500)
      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
                       'experiment' = experiment, 'method' = tool, 'topN' = i)
    }))
    dt1$endpoint <- 'PFI'
    # do the same for overall survival
    fit <- fit_RSF(surv, M, ntree = 500, 'OS.time', 'OS')
    dt2 <- do.call(rbind, lapply(2:10, function(i) {
      vars <- names(head(sort(fit$importance, decreasing = T), i))
      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'OS', 'OS.time'), 
                                M[,vars], ntree = 500)
      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
                       'experiment' = experiment, 'method' = tool, 'topN' = i)
    }))
    dt2$endpoint <- 'OS'
    return(rbind(dt1, dt2))
  }))
}))
parallel::stopCluster(cl)

@@

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 
                                        'experiments', 'settings'))
prog.clust <- pbapply::pbsapply(cl = cl, simplify = F, experiments, 
                                function(experiment){
    require(data.table)
    require(survminer)
    source(settings$utility_script)
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))

>>>>

---
title: "Manuscript Figures"
output:
  word_document: default
  html_document: default
params:
  settings_file: ''
  figures_folder: '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas/test'
  workdir: '.'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = params$workdir)
settings <- yaml::read_yaml(params$settings_file)
FigureFolder <- params$figures_folder
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
library(ggridges)
library(ggcorrplot)
library(ComplexHeatmap)
library(pheatmap)
theme_set(ggpubr::theme_pubclean())
geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'

@@

TOOLS <- c('maui_factors', 'mofa_factors', 'mcia_factors', 'pca_factors')

@@

# do unsupervised clustering experiment and compute AMI 
# downsample pan-cancer dataset 
df <- data.frame('id' = rownames(LFs$pca_factors$pancancer.cnv_gex_meth_mut))
df$pr <- surv[match(df$id, bcr_patient_barcode)]$project
ids <- lapply(split(df, df$pr), function(x) {
  if(nrow(x) > 200) {
    sample(x$id, 200)
  }
})
ids <- ids[lengths(ids) == 200]
k <- 5  
# sample cancer types and compute ami for different tools 
sampled <- sapply(simplify = F, 1:50, function(i) {
  unique(unlist(ids[sample(1:length(ids), k)]))
})
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('ids', 'df', 'LFs', 'settings', 'sampled', 'k'))
res <- do.call(rbind, pbapply::pblapply(cl = cl, 1:length(sampled), function(i) {
  require(data.table)
  source(settings$utility_script)
  selected <- sampled[[i]]
  ct <- df[match(selected, df$id),]$pr
  message("selected cancers:",paste0(unique(ct), collapse = ','))
  do.call(rbind, lapply(names(LFs), function(tool) {
    M <- scale(LFs[[tool]]$pancancer.cnv_gex_meth_mut[selected,])
    if(tool == 'maui_factors') {
      M <- scale(M)
    }
    ami <- compute_AMI(M, k, ct)
    return(data.table('i' = i, 'ami' = ami, 'method' = tool, 
                      'pr' = paste0(unique(ct), collapse = ',')))
  }))
}))
parallel::stopCluster(cl)
res$method <- toupper(gsub("_factors", "", res$method))

comps <- lapply(setdiff(unique(res$method), 'MAUI'), function(x) c('MAUI', x))
p5 <- ggboxplot(res, x = 'method', y = 'ami', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE,
                     comparisons = comps) +
  labs(y = "Adjusted Mutual Information", x = '') +
  theme(legend.position = 'none')

@@

multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
cluster.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  stats <- do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, 
                                            intersect(names(LFs[[tool]]), experiments), 
                    function(experiment) {
    message(experiment)
    source(settings$utility_script)
    library(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      #M <- remove_redundant_variables(M, perc = 99) 
      M <- scale(M)
    }

@@

# survival prediction accuracy benchmarking
# find survival-predictive LFs by univariate cox models and use them in RSF
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings', 'surv', 'clin_factors'))
surv.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
  message("survival benchmark for ",tool)
  do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source(settings$utility_script)
    require(data.table)
    require(survminer)
    M <- LFs[[tool]][[experiment]]
    clin <- process_clinical_covariates(rownames(M))
    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
    # compute pfi stats
    # fit a first model to filter variables 
    fit <- fit_RSF(surv, M, ntree = 500, 'PFI.time', 'PFI')
    # using a range of top variables, compute c index
    dt1 <- do.call(rbind, lapply(2:10, function(i) {
      vars <- names(head(sort(fit$importance, decreasing = T), i))
      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'PFI', 'PFI.time'), 
                                M[,vars], ntree = 500)
      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
                       'experiment' = experiment, 'method' = tool, 'topN' = i)
    }))
    dt1$endpoint <- 'PFI'
    # do the same for overall survival
    fit <- fit_RSF(surv, M, ntree = 500, 'OS.time', 'OS')
    dt2 <- do.call(rbind, lapply(2:10, function(i) {
      vars <- names(head(sort(fit$importance, decreasing = T), i))
      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'OS', 'OS.time'), 
                                M[,vars], ntree = 500)
      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
                       'experiment' = experiment, 'method' = tool, 'topN' = i)
    }))
    dt2$endpoint <- 'OS'
    return(rbind(dt1, dt2))
  }))
}))
parallel::stopCluster(cl)

@@

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 
                                        'experiments', 'settings'))
prog.clust <- pbapply::pbsapply(cl = cl, simplify = F, experiments, 
                                function(experiment){
    require(data.table)
    require(survminer)
    source(settings$utility_script)
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
